Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-11-15 paper/paper.tex
upated suppfile labels

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.
% Additionally, the use of the \newcommand function should be limited.

@@

\newfloat{suppfile}{thp}{lofsupfile}
%\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
\floatname{suppfile}{Supplementary file}

@@

\begin{abstract}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity~\citep{russell2018extreme}.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
\end{abstract}

@@

\figsupp[Number of PacBio CCSs and PCR strand exchange rate.]
{
The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
Note that these sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in Supplementary~file~\ref{suppfile:pacbio_analysis}.)
Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences derived from the same cDNA, and cell barcodes were used to collapse duplicate sequences from the same cell.
The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \FIGSUPP[genotypes]{StrandExchange} for more explanation).
From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
}
{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
\label{figsupp:CCSs}

@@

\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} for at detailed schematic).
However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
We therefore needed to enrich for the viral molecules, a process made challenging by the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end (the primer at the 5' end can be specific, but the one at the 3' end must bind the common 10X adaptor shared among all molecules).

@@

Because were were concerned that the multiplex PCR would still result in highly uneven amplification of different influenza cDNAs due to differences in their expression levels and lengths, the product of this multiplex PCR was subjected to eight additional individual emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 3' primer, using 1 ng of material in each reaction.
The material from these eight segment-specific PCRs was then pooled with the goal of obtaining a equimolar ratio of segments, and sequenced on one SMRT Cell in a PacBio RS II and one SMRT Cell of a PacBio Sequel. 
Detailed results from the analysis of these first two sequencing runs is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and Supplementary~file~\ref{suppfile:pacbio_analysis}.
These results showed that although the PCRs substantially enriched for influenza molecules, the relative coverage of the different viral genes was still very uneven, with the longer genes (especially the polymerase genes) severely under-sampled.

To try to improve coverage of the polymerase genes, we produced two new sequencing pools: one consisting of the five shortest viral segments (HA, NP, NA, M, and NS) from the aforementioned segment-specific emulsion PCRs, and the other consisting of the three longer polymerase segments (PB2, PB1, and PA).
The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\citep{xue2016propagation} and are presumably preferentially amplified during PCR.

To obtain more reads for longer full-length polymerase variants, we therefore subjected 10 ng of our amplified material for each polymerase segment to a bead selection using SPRIselect beads at a volume ratio of 0.4 to select for larger molecules. 
This selection removes most low-molecular weight DNA species including internally-deleted defective segments.
Material from this selection was then amplified using 16 (PB1) or 14 (PB2 and PA) cycles of a non-emulsion PCR using the standard conditions recommended by the Kapa HiFi Hotstart ReadyMix (extension at 67$^{\circ}$C for 2 minutes 15 seconds, and melting at 98$^{\circ}C$).
The use of relatively few PCR cycles was designed to prevent the occurrence of the artifacts (including strand exchange) that occur in non-emulsion PCRs as the reactions approach saturation.
We pooled the products of these reactions from this size-selection and sequenced on a SMRT Cell of a PacBio Sequel.
As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), this sequencing yielded modestly more full-length polymerase variants, but they were still heavily undersampled compared to other viral genes.

To further to improve recovery of full-length PB1, PB2, and PA, we therefore took an alternative approach that allowed us to perform a specific PCR for full-length polymerase variants.
Specifically, we circularized the template molecules (including the full length of genes plus the cell barcode, UMI, and 10X adaptor), and then used two segment-specific primers that annealed in apposition near the center of each polymerase gene to linearize these circular molecules.
Only molecules that contain the middle of the polymerase genes (which are typically full-length) are linearized by this process.
In the downstream computational analysis, we can then determine the full sequence of the gene as well as the cell barcode of the initial molecule from which the linearized molecule is derived.
Specifically, we first used 2.5 ng of our already-amplified segment-specific material in a 10-cycle PCR to append circularization adapters (see \FIGDATA[genotypes]{primer_sequences} for sequences), and cleaned the resultant mixture using SPRIselect beads at a volume ratio of 0.4.
We then used 10 ng of this amplified material in a 20$\mu$l NEBuilder reaction using an extended reaction time of 50 minutes in order to circularize the molecules.
We next incubated these reactions for 1 hour at 37$^{\circ}$C with exonuclease V and additional ATP to a final increase in concentration of 1 mM to digest all non-circularized molecules.
The circularized and digested material was then cleaned using SPRIselect beads at a volume ratio of 0.4.
This material was then used as template for three non-emulsion PCRs specific to PB2, PB1, or PA, using two segment-specific primers that align to the central portion of each gene but in apposition to each another (see \FIGDATA[genotypes]{primer_sequences} for sequences).
These linearization reactions used 20 (PB2) or 26 (PB1 and PA) PCR cycles, and the resulting products were cleaned using SPRIselect beads at a volume ratio of 1.0.
This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRT Cell of PacBio Sequel. 
As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), this process efficiently yielded many full-length polymerase variants.

The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[genotypes]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[genotypes]{StrandExchange} for an illustration of how this is determined).
A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}).
Importantly, the various PCR biases and enrichment schemes mean that the coverage of molecules by the PacBio sequencing is not proportional to their original abundance in the starting mRNA.
However, as described in the computational analysis section below, the final analyses use the cell barcodes and UMIs in conjunction with the standard 10X Illumina sequencing to ensure that none of the conclusions are affected by the disproportionate amplification of some molecules during the PacBio library preparation (for instance, duplicate UMIs are removed from the PacBio data, and all conclusions about gene abundance or absence are based on the Illumina data).

@@

The PacBio sequences of the full-length viral genes were analyzed as follows.
First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
We further processed these CCSs using custom Python code and the \texttt{minimap2}~\citep{li2018minimap2} long-read aligner (version 2.11-r797).
The Python code has been implemented in the API of \texttt{dms\_tools2}~\citep[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:pacbio_analysis}.
We refer the reader to this notebook for a detailed description and extensive plots showing the results at each step.
Here is a brief summary: we filtered for CCSs that had the expected 5' termini (from the influenza-specific primers) and 3' termini (corresponding to the 10X adaptor), and for which we could identify the cell barcode, UMI, and polyA tail.
We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\FIGSUPP[genotypes]{StrandExchange}), and reassuringly found that strand exchange was rare (\FIGSUPP[genotypes]{CCSs}).
The small number of CCSs with identifiable strand exchange were filtered from further analysis.
We then further filtered for CCSs that contained valid cell barcodes as identified by the \texttt{cellranger} pipeline, and kept just one CCS per UMI (preferentially retaining high-quality CCSs that aligned to full-length cDNAs).
We then removed from the CCSs the barcoding synonymous mutations that we had engineered into one of the two viral variants.
Finally, we used the CCSs to call the sequence of the viral gene in each cell, calling mutations separately for each viral barcode variant.
We called mutations (insertions, deletions, and substitutions) in the viral gene sequences as follows:
\begin{enumerate}
\item Mutations with accuracies less than 0.999 (which constitute $<$0.5\% of all mutations) were ignored.
\item If all CCSs for a particular viral-barcode variant of a gene in a cell were wild-type, it was called as wild type.
\item If any CCSs for a particular viral-barcode variant of gene in a cell had a mutation, then require at least two CCSs to call the sequence.
\item If at least two and $>$30\% of the CCSs had a specific mutation, then call that mutation as present and note its frequency among the CCSs. The exception was single-nucleotide indels in homopolymers, for which we required three CCSs to call a mutation (the reason is that the main mode of PacBio sequencing errors is short indels in homopolymers).
\end{enumerate}
The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
We could call the sequences of all expressed viral genes in about half of the infected cells (\FIGSUPP[genotypes]{ncells}).
The mutations called using this pipeline are shown in \FIG{genotypes}, and \FIGDATA[genotypes]{genotypes} gives the number of CCSs supporting each mutation call.
The called sequences of the viral genes were added to the annotated cell-gene matrix.

Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
This analysis utilized a variety of R and Bioconductor~\citep{huber2015orchestrating} packages, including \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:monocle_analysis}.
We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{transcriptomics}B.
We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \FIGSUPP[transcriptomics]{frac_has_gene}). 
To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}, respectively.
We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).

>>>>

\documentclass[9pt,lineno]{template}

@@

\newfloat{suppfile}{thp}{lofsupfile}
\floatname{suppfile}{Supplementary File}

@@

\begin{abstract}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
\end{abstract}

@@

\figsupp[Number of PacBio CCSs and PCR strand exchange rate.]
{
The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
Note that these sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in \SUPPFILE{pacbio_analysis}.)
Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences derived from the same cDNA, and cell barcodes were used to collapse duplicate sequences from the same cell.
The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \FIGSUPP[genotypes]{StrandExchange} for more explanation).
From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
}
{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
\label{figsupp:CCSs}

@@

\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} for at detailed schematic).
However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
We therefore needed to enrich for the viral molecules, a process made challenging by the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end (the primer at the 5' end can be specific, but the one at the 3' end must bind the common 10X adaptor shared among all molecules).

@@

Because were were concerned that the multiplex PCR would still result in highly uneven amplification of different influenza cDNAs due to differences in their expression levels and lengths, the product of this multiplex PCR was subjected to eight additional individual emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 3' primer, using 1 ng of material in each reaction.
The material from these eight segment-specific PCRs was then pooled with the goal of obtaining a equimolar ratio of segments, and sequenced on one SMRT Cell in a PacBio RS II and one SMRT Cell of a PacBio Sequel. 
Detailed results from the analysis of these first two sequencing runs is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and \SUPPFILE{pacbio_analysis}.
These results showed that although the PCRs substantially enriched for influenza molecules, the relative coverage of the different viral genes was still very uneven, with the longer genes (especially the polymerase genes) severely under-sampled.

To try to improve coverage of the polymerase genes, we produced two new sequencing pools: one consisting of the five shortest viral segments (HA, NP, NA, M, and NS) from the aforementioned segment-specific emulsion PCRs, and the other consisting of the three longer polymerase segments (PB2, PB1, and PA).
The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\citep{xue2016propagation} and are presumably preferentially amplified during PCR.

To obtain more reads for longer full-length polymerase variants, we therefore subjected 10 ng of our amplified material for each polymerase segment to a bead selection using SPRIselect beads at a volume ratio of 0.4 to select for larger molecules. 
This selection removes most low-molecular weight DNA species including internally-deleted defective segments.
Material from this selection was then amplified using 16 (PB1) or 14 (PB2 and PA) cycles of a non-emulsion PCR using the standard conditions recommended by the Kapa HiFi Hotstart ReadyMix (extension at 67$^{\circ}$C for 2 minutes 15 seconds, and melting at 98$^{\circ}C$).
The use of relatively few PCR cycles was designed to prevent the occurrence of the artifacts (including strand exchange) that occur in non-emulsion PCRs as the reactions approach saturation.
We pooled the products of these reactions from this size-selection and sequenced on a SMRT Cell of a PacBio Sequel.
As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), this sequencing yielded modestly more full-length polymerase variants, but they were still heavily undersampled compared to other viral genes.

To further to improve recovery of full-length PB1, PB2, and PA, we therefore took an alternative approach that allowed us to perform a specific PCR for full-length polymerase variants.
Specifically, we circularized the template molecules (including the full length of genes plus the cell barcode, UMI, and 10X adaptor), and then used two segment-specific primers that annealed in apposition near the center of each polymerase gene to linearize these circular molecules.
Only molecules that contain the middle of the polymerase genes (which are typically full-length) are linearized by this process.
In the downstream computational analysis, we can then determine the full sequence of the gene as well as the cell barcode of the initial molecule from which the linearized molecule is derived.
Specifically, we first used 2.5 ng of our already-amplified segment-specific material in a 10-cycle PCR to append circularization adapters (see \FIGDATA[genotypes]{primer_sequences} for sequences), and cleaned the resultant mixture using SPRIselect beads at a volume ratio of 0.4.
We then used 10 ng of this amplified material in a 20$\mu$l NEBuilder reaction using an extended reaction time of 50 minutes in order to circularize the molecules.
We next incubated these reactions for 1 hour at 37$^{\circ}$C with exonuclease V and additional ATP to a final increase in concentration of 1 mM to digest all non-circularized molecules.
The circularized and digested material was then cleaned using SPRIselect beads at a volume ratio of 0.4.
This material was then used as template for three non-emulsion PCRs specific to PB2, PB1, or PA, using two segment-specific primers that align to the central portion of each gene but in apposition to each another (see \FIGDATA[genotypes]{primer_sequences} for sequences).
These linearization reactions used 20 (PB2) or 26 (PB1 and PA) PCR cycles, and the resulting products were cleaned using SPRIselect beads at a volume ratio of 1.0.
This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRT Cell of PacBio Sequel. 
As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), this process efficiently yielded many full-length polymerase variants.

The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[genotypes]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[genotypes]{StrandExchange} for an illustration of how this is determined).
A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
Importantly, the various PCR biases and enrichment schemes mean that the coverage of molecules by the PacBio sequencing is not proportional to their original abundance in the starting mRNA.
However, as described in the computational analysis section below, the final analyses use the cell barcodes and UMIs in conjunction with the standard 10X Illumina sequencing to ensure that none of the conclusions are affected by the disproportionate amplification of some molecules during the PacBio library preparation (for instance, duplicate UMIs are removed from the PacBio data, and all conclusions about gene abundance or absence are based on the Illumina data).

@@

The PacBio sequences of the full-length viral genes were analyzed as follows.
First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
We further processed these CCSs using custom Python code and the \texttt{minimap2}~\citep{li2018minimap2} long-read aligner (version 2.11-r797).
The Python code has been implemented in the API of \texttt{dms\_tools2}~\citep[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{pacbio_analysis}.
We refer the reader to this notebook for a detailed description and extensive plots showing the results at each step.
Here is a brief summary: we filtered for CCSs that had the expected 5' termini (from the influenza-specific primers) and 3' termini (corresponding to the 10X adaptor), and for which we could identify the cell barcode, UMI, and polyA tail.
We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\FIGSUPP[genotypes]{StrandExchange}), and reassuringly found that strand exchange was rare (\FIGSUPP[genotypes]{CCSs}).
The small number of CCSs with identifiable strand exchange were filtered from further analysis.
We then further filtered for CCSs that contained valid cell barcodes as identified by the \texttt{cellranger} pipeline, and kept just one CCS per UMI (preferentially retaining high-quality CCSs that aligned to full-length cDNAs).
We then removed from the CCSs the barcoding synonymous mutations that we had engineered into one of the two viral variants.
Finally, we used the CCSs to call the sequence of the viral gene in each cell, calling mutations separately for each viral barcode variant.
We called mutations (insertions, deletions, and substitutions) in the viral gene sequences as follows:
\begin{enumerate}
\item Mutations with accuracies less than 0.999 (which constitute $<$0.5\% of all mutations) were ignored.
\item If all CCSs for a particular viral-barcode variant of a gene in a cell were wild-type, it was called as wild type.
\item If any CCSs for a particular viral-barcode variant of gene in a cell had a mutation, then require at least two CCSs to call the sequence.
\item If at least two and $>$30\% of the CCSs had a specific mutation, then call that mutation as present and note its frequency among the CCSs. The exception was single-nucleotide indels in homopolymers, for which we required three CCSs to call a mutation (the reason is that the main mode of PacBio sequencing errors is short indels in homopolymers).
\end{enumerate}
The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
We could call the sequences of all expressed viral genes in about half of the infected cells (\FIGSUPP[genotypes]{ncells}).
The mutations called using this pipeline are shown in \FIG{genotypes}, and \FIGDATA[genotypes]{genotypes} gives the number of CCSs supporting each mutation call.
The called sequences of the viral genes were added to the annotated cell-gene matrix.

Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
This analysis utilized a variety of R and Bioconductor~\citep{huber2015orchestrating} packages, including \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{monocle_analysis}.
We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{transcriptomics}B.
We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \FIGSUPP[transcriptomics]{frac_has_gene}). 
To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}, respectively.
We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).
